Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-06-16 latex/intro.tex
Working towards revision. Modified intro, now working on refactoring code

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. Selecting the optimal set
of measurements holds particular significance when inferring
parameters of a physical process. Unlike many other fields, where
measurements are fixed, in this context, we have the freedom to choose
which measurements to take. This freedom should be taken advantage of:
measurements should be chosen to enhance accuracy, reduce costs, or
both. Whether measurements involve specifying electrode locations on
the skin in electric impedance tomography \cite{horesh2010impedance},
determining certain wavelengths in MRI \cite{horesh2008mri},
positioning sensors for detecting ground-reflected waves in the search
for oil \cite{horesh2008borehole}, or optimizing sensor locations in a
seismographic network \cite{rabinowitz1990, steinberg1995} --- the
selection of optimal measurements, referred to as the problem of
\emph{optimal design}, becomes crucial. Optimal measurements are
chosen according to specific \emph{design criteria}, with A- and
D-optimality being two of the most widely recognized and extensively
studied design criteria \cite{Chaloner1995,
  AlexanderianGloorGhattas14}.

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  nyberg2012, fedorov1997, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy problem: inferring
the initial condition of the 1D heat equation (details in the
supplementary material). In Fig.~\ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. We refer to this intriguing
phenomenon as \emph{measurement clusterization} \cite{Ucinski05}, and
we refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.pdf}
    \caption{Measurement clusterization for optimal designs when
      inverting for the initial condition of the 1D heat equation (see
      supplementary material for details). Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid with the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging close measurements
\cite{fedorov1997}; however, this strategy merely overlooks the
phenomenon of measurement clusterization. An alternative solution lies
in \emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{nyberg2012}.

@@

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties for posterior
covariance eigenvectors in observation space. A similar conclusion was
reached by Koval et al.~\cite{koval2020}, who showed that A-optimal
designs are best constructed in the space of observations.

Before we state Theorem \ref{thm:char} we give some definitions: Let
\(\hilp, \hilo\) Hilbert spaces, \(\fwd:\hilp \to \hilo\) a linear
compact operator. Let \(\obs: \hilo \to \mathbb{R}^m\) a linear
measurement operator, where \(m \in \mathbb{N}\) is the number of
measurements taken. Let \(\sigma^2 \in \mathbb{R}_{+}\) observation
noise variance, \(\data = \obs \fwd \param + \eps\), where \(\eps \in
\mathbb{R}^m\) is iid \(\mathcal{N}(0, \sigma^2)\) noise. Let \(\pr
\sim \mathcal{N}(\prmean, \prcov)\) prior Gaussian measure on
\(\hilp\), where \(\prcov\) is the prior covariance operator and let
\(\post\) the posterior (Gaussian) measure.

\begin{restatable}{theorem}{char}\label{thm:char}
  Let:
  \begin{itemize}
    \item The D-optimality design criterion
    \cite{AlexanderianGloorGhattas14}:
    \begin{align*}
      \begin{split}
        \tar(\obs) %:&= \mathbb{E}_{\data}\left [ D_{\text{KL}} (\post || \pr ) \right ] \\
        % 
        % 
        % 
        &= \frac12 \log \det ( I + \sigma^{-2} \prcov^{1/2} \fwd ^*
        \obs^* \obs \fwd \prcov^{1/2}), 
      \end{split}
    \end{align*}
  \item \(\obs\) a D-optimal design operator
    \begin{equation*}
      \obs = \argmax_{\|\meas_j\| = 1, j=1,\dots,m}\tar(\obs),
    \end{equation*}
  \item \(\{\lambda_i\}_{i=1}^\infty\) eigenvalues of
    \(\fwd\prcov\fwd^*\) in decreasing order of magnitude.
    %% \item \(\{\ev_i\}_{i=1}^\infty\) their corresponding eigenvectors.
  \item \(\{\eta_i\}_{i=1}^\infty\) eigenvalues of \(\obs^*\obs\).

  \end{itemize}

  Then:
  \begin{enumerate}
  \item  \(\tr{\obs^*\obs} = m\).
  \item \(\obs^*\obs\) and \(\fwd\prcov\fwd^*\) are simultaneously
    diagonalizable.
  \item \(k := \rank \obs^*\obs \leq m\) and
    \begin{equation*}
      \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). %= \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
    \end{equation*}
  %% \item 
  %%   \begin{equation*}
  %%     k = \argmax \left \{ k:\lambda_k^{-1} < \sigma^{-2}\frac{m}{k} + \frac{1}{k} \sum_{j=1}^{k}
  %%     \lambda_j^{-1} \right \}.
  %%   \end{equation*}
  \item
    \begin{equation*}
        \eta_i = \begin{cases}
          \frac{m}{k} - \sigma^2 \lambda_i^{-1} + \sigma^2 \frac{1}{k} \sum_{j=1}^k \lambda_j^{-1} & 1 \leq i \leq k \\
          0 & i > k 
        \end{cases}.
    \end{equation*}
  \item The covariance of the pushforwad \(\fwd_{*} \post\) is \(\left
    ( (\fwd \prcov \fwd^*)^{-1} + \sigma^{-2} \obs^*\obs \right
    )^{-1}\) and its eigenvalues are
    \begin{equation*}
      \theta_i =
      \begin{cases}
        \left(\frac{\sum_{j=1}^k \lambda_j^{-1} + \sigma^{-2}m}{k} \right )^{-1} & i \leq k \\
        \lambda_i &  i > k 
      \end{cases}
    \end{equation*}
  \end{enumerate}
%\end{theorem}
\end{restatable}

In the process of proving Theorem \ref{thm:char} we prove and
generalize several lemmas. Among those, is Lemma \ref{lemma:free},
which is (to the the best of my knowledge) a novel lemma in linear
algebra: We decompose a symmetric positive definite matrix \(M \in
\mathbb{R}^{k \times k}\) with \(\ttr M = m \in \mathbb{N}\) as \(M =
AA^t\), where \(A\) has unit norm columns.

@@

>>>>

\section{Introduction}\label{section:intro}
The practice of experimental design entails specifying every detail of
running an experiment \cite{chaloner}: at its simplest form,
experimental design involves choosing a set of measurements to be
taken. In a medical experiment, experimental design may involve
choosing participants to whom some drug will be administered, and
measuring their blood pressure. Experimental design also arises 
process of searching for oil. Since measuremnts are expensive and
require digging holes in the ground, only a limited number of
measuremnts can be taken, hence carefully choosing measuremt locations
is crucial.

Specifiyng the right set of measurements holds particular significance
when solving an \emph{inverse problem} --- i.e. making inference of
the physical world utilizing a physical model of some phenomenon of
interest. In an inverse problem, we seek to infer physical quantities,
based on observations and a model of the physical world, typically
phrased in the language of ordinary or partial differential equations.

Inverse problems are ubiquitous in science and engineering, and we
will now give several illustrative examples. In electric impedance
tomography (EIT) we seek to infer the impedance of the inside of a
human body from observations of electrical current impedence measured
at specific electrode locations on the skin, leveraging known
equations of electric current flow
\cite{horesh2010impedance}. Magnetic Resonance Imaging (MRI) also
entails solving an inverse problem, where radio-frequency pulses are
sent through the human body in the presence of a strong magnetic field
and the response of our bodies' contents to such perutrbations is
measured, allowing a radiologist to view the internals of a body in a
noninvasive manner \cite{horesh2008mri}. Yet another example of an
inverse problem arises in oil and gas exploration, where acoustic
waves are sent through "boreholes" --- deep cylindrical wells drilled
into the ground --- and data of travel and return times of said waves
are recorded. Since travel and return time is influenced by the
properties of the subsurface materials, such as density, elasticity,
and the presence of fluids or voids, we can combine travel data with a
geophisical models of contents of the earth's crust to facilitate
reconstructing and inferring the structure of the subsurface in a
process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola}. Importantly, note that in the
above examples our goal is to infer some function $f \in C(\Omega)$,
where \(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain
of interest.

%% In EIT, electrode locations determine the quality of the
%% reconstructed impedance field, while in MRI utilizng certain
%% radio-frequency pulses may give more accurate information than
%% others. In borehole tomogrpahy, only several boreholes could be dug
%% and the choice of their locations is also crucial for oil and gas
%% discovery. Thus, the selection of optimal measurements, referred to as
%% the problem of \emph{optimal design}, becomes crucial.
%% in a specific location (in a typical
%% medical application to measure inverventions on, . These experimental
%% units may be individual patients, but also Although common sense is
%% useful in this process, it is often the case that naive intuition
%% results in experiments that do not generate the maximal amount of
%% information possible. To this end

We will now introduce a toy model which we will use to define a toy
inverse problem: the heat equation in one dimension with a homogeneous
Dirichlet boundary condition. The equations are presented below for
completeness, but readers less familiar with partial differential
equations should think of the heat equation as follows: before us
there is a perfectly insulated metal rod. At each point on the rod the
temperature is different, and we call this temperature distribution
the \emph{initial condition}. We wait for a short time and let the
heat dissipate a bit inside the rod. The final heat distribution in
the rod is dictated by the \emph{1D heat equation}:
\begin{subequations}\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty),\\ u
      &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty),\\ u &= u_0
        &&\qquad \text{on }[0,1] \times \{0\}.
  \end{alignat}
\end{subequations}

In the inverse problem, our goal is to infer the initial condition
(i.e.~initial temperature distribution inside the rod). We are allowed
to take some number of temperature readings on the rod once our
waiting time has passed. The inverse problem of the 1D heat equation
is difficult ("ill-posed") since heat spreads in a diffusive manner
and any roughness in the initial condition is quickly smoothed, see
Supplementary movies S1 and S2. We will give a fully Bayesian
formulation of this inverse problem in Section \ref{section:prelims}.

In all of the examples of inverse problems we have seen, as well as in
many other applications, measurements should be chosen to most benefit
the user. Hence, the user may consider some utility, called a
\emph{design criterion} and take measurements that maximize this
utility. Two of the most widely recognized and extensively studied
design criteria are A- and D-optimality. We focus on the Bayesian
D-optimality criterion, which carries a simple and intuitive meaning:
measurements chosen according to the D-optimality criterion maximize
the expected Kullback-Liebler divergence from posterior to prior
\cite{Chaloner1995, AlexanderianGloorGhattas14}, see also precise
definition in Section \ref{subsec:D optimal design}. We refer to a set
of measuremnts that maximizes the Bayesian D-optimality criterion a
\emph{D-optimal design}.

Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}. For our purposes, it suffices for
measurements to be visibly close for clusterization to hold. We refer
to a design that exhibits measurement clusterization as a
\emph{clustered design},

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.pdf}
    \caption{Measurement clusterization for D-optimal designs for the
      inverse problem of the heat equation. Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid on the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

Clusterization should not be confused with either repetition nor with
replication, which are commonly viewed as beneficial and even
necessary aspects of an optimal design \cite{morris, fisher
  schafer2001replication}. For example, in his famous milk and tea
experiment, \cite[Section 11]{fisher} suggested that repetition is a
"way of enlarging the experiment and, thereby, increasing its
sensitiveness". In another example, \cite{fay2000rainfall} measured
the effect of rainfall on grass growth in plots of land. The
experiment involved fifteen "rainfall manipulation shelters", where
"Four rainfall manipulation treatments (three replicates) then were
assigned to 12 of the plots". It seems reasonable the researchers
replicate the phenomenon they are trying to study. Clusterization
seems like a whole different beast: a clustered design in the rainfall
experiment would imply the researchers should take a repeated
measurement \emph{on the same plot}, at the expense of measuring grass
growth in other plots. In sharp contrast to repetition and
replication, clusterization is highly nonintuitive and, as we will
see, its origins run considerably deeper.

Researchers of inverse problems widely agree that measurement
clusterization is undesirable \cite{fedorov1996, nyberg2012,
  fedorov1997, Ucinski05, neitzel2019sparse}, prompting the
exploration of various remedies to address this issue. One approach
involves merging close measurements \cite{fedorov1997}; however, this
strategy merely overlooks the phenomenon of measurement
clusterization. An alternative solution lies in
\emph{clusterization-free design}s, where measurement locations are
deliberately chosen to be distant from one another. This can be
achieved by imposing distance constraints between measurements or by
introducing correlated errors that account for both observation error
and model misspecification \cite{Ucinski05}. For instance, in the
context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{nyberg2012}.

@@

The first insight our study gives is that a D-optimal design aims to
uniformly reduce uncertainties not over space (i.e.~$\Omega$), but
rather over \emph{eigenvectors}. This insight is a part of Theorem
\ref{thm:char}, see Section \ref{section:vanishing} for precise
statement and proof, and see Fig.~\ref{cluste} for an illustration.

In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the sapce of \emph{observations}. Again, a precise
statement could be found in Section \ref{section:vanishing}.  This
result is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.

In the process of proving Theorem \ref{thm:char} we prove and
generalize several lemmas. Among those, is Lemma \ref{lemma:free},
which is a lemma in linear algebra that (to the the best of my
knowledge) is novel: We decompose a symmetric positive definite matrix
\(M \in \mathbb{R}^{k \times k}\) with \(\ttr M = m \in \mathbb{N}\)
as \(M = AA^t\), where \(A\in \mathbb{R}^{k \times m}\) has unit norm
columns.

@@

Finally, we must stress that we do not suggest that clustered designs
should be abandoned, nor do we support the use of clustered designs in
any circumstances. We merely investigate an intriguing phenomenon of
clusterization and seek a deeper understanding for it. Researchers in
other disciplinesis not viewed as a problem. For example, \cite{lozan}
suggest in the introduction section that repeated measurements are
necessary. Our results align with that common wisdom and settle the
seemed contradiction.

### 2024-06-16 latex/numerix.tex
Working towards revision. Modified intro, now working on refactoring code

%\documentclass{siamonline220329}
\documentclass{amsart}
%%\documentclass{article}
\input{definitions}

\title[Supplementary Material]{Supplementary Material for: Measurement
  Clusterization in D-optimal Designs for Bayesian Linear Inverse
  Problems over Hilbert Spaces}

\begin{document}
\maketitle

\section{The inverse problem of the 1D heat equation}\label{section:example}
The mathematical foundations of inversion in function spaces can be
found in \cite{Stuart10}. Our goal is to infer the initial condition of
the one-dimensional heat equation with a homogeneous Dirichlet
boundary condition:
\begin{subequations}\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty),\\
      u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty),\\
        u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}.
  \end{alignat}
\end{subequations}

Model the prior for the initial condition as $u_0 \sim
\normal(0,\prcov)$, for $\prcov = (-\Delta)^{-1}$ with a homogeneous
Dirichlet boundary condition. Let the final time $T = 0.03$ and denote
$\fwd$ the forward operator, so that $u(\cdot, T) = \fwd u_0$. Let
$\obs$ a measurement operator, so that $u(\x_j,T) = (\obs u)_j,
j=1,\dots,m$, where $m$ is the number of allowed measurements. Assume
iid centered Gaussian observation error, so observations are given by
$u(\x_j,T) + \eps(\x_j)$ with $\eps(\x_j) \sim \normal(0, \sigma^2)$
iid, $\sigma = 0.05$. Measurement locations $\x_j,j=1,\dots,m$ are
chosen according to the D-optimality criterion of
\cite{AlexanderianGloorGhattas14}.

Implementation of this inverse problem can be found in the
repository \url{https://github.com/yairdaon/OED}, with an optimization
routine for finding D-optimal designs. Unfortunately, I was not able
to install PyOED \cite{attia2023pyoed}, nor was I able to utilize
hIPPYlib \cite{VillaPetraGhattas16, VillaPetraGhattas18,
VillaPetraGhattas21}, hence the inverse problem is implemented from
scratch.

\section{Generalizations of known lemmas}

The following lemma is generalized from \cite[Chapter 9, Theorem 4,
p. 127]{Lax07}
\begin{lemma}\label{lemma:lax}
  Let $Y(t)$ be a differentiable operator-valued function. Assume 
  $I+Y(t)$ is invertible, $Y(t)$ self-adjoint and trace-class. Then
  \begin{equation*}
    \frac{\der \log \det (I+Y(t))}{\der t} = \tr{(I+Y(t))^{-1} \dot{Y}(t)}.
  \end{equation*}
\end{lemma}

\begin{proof}
  Consider a differentiable operator-valued function $X(t)$ such that
  $X(0) = 0$ and $X(t)$ is positive, self-adjoint and trace-class for
  every $t\in \R$. We denote the eigenvalues of this operator by
  $\lambda_k(X(t))$ and sometimes drop the dependence on $X(t)$, so
  $\lambda_k = \lambda_k(X(t))$.  Then $\det (I+X(t))
  = \prod_{k=1}^{\infty} (1+\lambda_k) < \infty$ and this is finite by
  the arguments given in \cite{AlexanderianGloorGhattas14}. The full
  derivative is \begin{align*} \frac{\der \det (I+X(t))}{\der t}
    % 
    % 
    % 
    &= \sum_{k=1}^{\infty} 
    \frac{\partial \det (I+X(s))}{\partial (1+\lambda_k)}\Big |_{s=t}
    \frac{\der (1+\lambda_k)}{\der t} \\
    % 
    %
    %
    &= \sum_{k=1}^{\infty} \frac{\partial \prod_{l=1}^{\infty}
      (1+\lambda_l(s))}{\partial (1+\lambda_k)}\Big |_{s=t}
    \frac{\der (1+\lambda_k)}{\der t} \\
    %
    %
    %
    &= \sum_{k=1}^{\infty} \prod_{l=1, l\neq k}^{\infty}
      (1+\lambda_l(s)) \frac{\partial (1+\lambda_k(s))}{\partial (1+\lambda_k)}\Big |_{s=t}
    \frac{\der (1+\lambda_k)}{\der t} \\
    %
    %
    %    
    &= \sum_{k=1}^{\infty} \frac{\prod_{l=1}^{\infty}
      (1+\lambda_l(s))}{(1+\lambda_k)}\Big |_{s=t}
    \dot{\lambda_k}(X(t)) \\
    % 
    % 
    % 
    &= \sum_{k=1}^{\infty} \frac{\det (I+X(t))}{1 +\lambda_k} \dot{\lambda_k}(X(t)).
  \end{align*}
  The assumption $X(0) = 0$ means $\lambda_k(X(0)) = 0,\ \forall k \geq 1$. Thus:
  \begin{align*}
    \frac{\der (I+\det X(t))}{\der t}\Big |_{t=0} 
    = \sum_{k=1}^{\infty} \dot{\lambda_k}(X(0)) 
    = \frac{\der }{\der t}\tr{X(0)}
    = \tr{\dot{X}(0)},
  \end{align*}
  where the second equality follows by monotone convergence. 
  Let $Y(t)$ a trace-class self-adjoint operator such that 
  $I+Y(t)$ is invertible.
  Define $X(t)$ via $I+X(t) = (I+Y(0))^{-1/2} (I+Y(t)) (I+Y(0))^{-1/2}$. 
  We show $X(t)$ satisfies the conditions above. It is trace-class:
  \begin{align*}
    \tr{X(t)} = \tr{(I+Y(0))^{-1} (I+Y(t)) - I}
    \leq \tr{I+Y(t) - I}< \infty,
  \end{align*}
  since $Y(t)$ is trace-class. It is also clear that
  $X(0) = 0$ and $X(t)$ is self-adjoint.
  $I+Y(t) = (I+Y(0))^{1/2}(I+X(t))(I+Y(0))^{1/2}$, so
  \begin{align*}
    \frac{\der \det (I+Y(t))}{\der t}|_{t=0} 
    &= \det (I+Y(0))\frac{\der \det (I+X(t))}{\der t}\Big |_{t=0} \\
    % 
    % 
    % 
    &= \det (I+Y(0)) \tr{\dot{X}(0)} \\
    % 
    % 
    % 
    &= \det (I+Y(0)) \tr{(I+Y(0))^{-1} \dot{Y}(0)}.
  \end{align*}
  Consequently, by the one-variable chain rule:
  \begin{align*}
    \frac{\der \log \det (I+Y(t))}{\der t}\Big |_{t=0} &=
    % 
    % 
    % 
    \frac{1}{\det (I+Y(0))}\frac{\der \det (I+Y(t))}{\der t}\Big |_{t=0} \\ 
    % 
    % 
    % 
    &= \tr{ (I+Y(t))^{-1} \dot{Y}(t)} \big |_{t=0}.
  \end{align*}
  There is nothing special about $t_0 = 0$ --- we could have chosen
  any other $t_0$ instead. Thus, the relation holds for all $t$.
\end{proof}

The following is a generalization of the Matrix Determinant Lemma to
Hilbert spaces.
\begin{lemma}%\label{lemma:MDL}
  Let $\hil$ a separable Hilbert space, $u,v\in \hil$ and $A: \hil \to
  \hil$ an invertible linear operator such that $\tr{A-I} <
  \infty$. Then $\det A$ and $\det A + uv^*$ are well defined and
  \begin{equation*}
    \det (A + uv^*) = (1 + \langle A^{-1} u, v \rangle ) \det A,
  \end{equation*}
  where $(A + uv^*)w := Aw + \langle v,w \rangle u$.
\end{lemma}
\begin{proof}
  In this proof we rely on definitions and results from
  \cite{simon1977}. First, consider $B := I + xy^*$ for some $x,y \in
  \hil$. We construct an eigenbasis for $B$ and use that to show $\det
  B = 1 + \langle x, y \rangle$. First let $x_1 := x$.  Now, if $x
  \parallel y$, take $\{x_n \}_{n=2}^{\infty}$ an orthogonal basis for
  $span\{x_1\} ^{\perp}$. If, on the other hand, $x \not \parallel y$, let
  \begin{equation*}
    x_2 := x - \frac{ \langle x, y\rangle}{\|y\|^2}y
  \end{equation*}
  and it is easy to verify that $x_2 \perp y$ and $span \{x,y\} = span
  \{x_1,x_2\}$. Take $\{x_n \}_{n=3}^{\infty}$ an orthogonal basis for
  $span\{x_1,x_2\} ^{\perp}$. In both cases,
  \begin{equation*}
    B x_n =
    \begin{cases}
      (1 + \langle x, y \rangle) x_n & n = 1 \\
      x_n                            & n \neq 1,
    \end{cases}
  \end{equation*}
  and so $\det B = 1 + \langle x, y \rangle$.

  It is easy to verify that $uv^*$ is trace-class and since $\tr{A-I}
  < \infty$, also $\tr{A + uv^* - I} < \infty$ (sum of two trace-class
  operators is trace-class). Thus $\det A$ and $\det (A+uv^*)$ are
  well defined. Let $x:=A^{-1}u$ and $y := v$:
  \begin{equation*}
    \det (A + uv^*) = \det A \ \det(I+A^{-1}uv^*) =
    (1 + \langle A^{-1}u, v \rangle) \det A .
  \end{equation*}
\end{proof}

%\bibliographystyle{siamplain}
\bibliographystyle{amsplain}
\bibliography{/home/yair/projects/bibtex.bib}

\end{document}

>>>>

\section{Numerical Experiments}
We provide numerical evidence supporting our conclusions.

\subsection{Heat Equation}
Our goal is to infer the initial condition of
the one-dimensional heat equation with a homogeneous Dirichlet
boundary condition:
\begin{subequations}\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty),\\
      u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty),\\
        u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}.
  \end{alignat}
\end{subequations}

Model the prior for the initial condition as $u_0 \sim
\normal(0,\prcov)$, for $\prcov = (-\Delta)^{-1}$ with a homogeneous
Dirichlet boundary condition. Let the final time $T = 0.03$ and denote
$\fwd$ the forward operator, so that $u(\cdot, T) = \fwd u_0$. Let
$\obs$ a measurement operator, so that $u(\x_j,T) = (\obs u)_j,
j=1,\dots,m$, where $m$ is the number of allowed measurements. Assume
iid centered Gaussian observation error, so observations are given by
$u(\x_j,T) + \eps(\x_j)$ with $\eps(\x_j) \sim \normal(0, \sigma^2)$
iid, $\sigma = 0.05$. Measurement locations $\x_j,j=1,\dots,m$ are
chosen according to the D-optimality criterion
\cite{AlexanderianGloorGhattas14}.

Implementation of this inverse problem can be found in this 
\href{https://github.com/yairdaon/OED}{repository}, including an optimization
routine for finding D-optimal designs. Unfortunately, I was not able
to install PyOED \cite{attia2023pyoed}, nor was I able to utilize
hIPPYlib \cite{VillaPetraGhattas16, VillaPetraGhattas18,
  VillaPetraGhattas21}, hence the inverse problem is implemented from
scratch.

\subsection{Simulating Lemma ... }
In theorem 17 we explicitly construct a D-optimal design in a Hilbert
space where the base vectors diagonalize $\fwd \prcov\fwd^*$. We
explicitly implement the algorithm outlined in Lemma

\subsection{Correlated errors}
We simulate the 1D heat eq

### 2024-06-16 latex/prelims.tex
Working towards revision. Modified intro, now working on refactoring code

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let \(\hilp\) and \(\hilo\) separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively), and let
\(\fwd: \hilp \to \hilo\) the \emph{forward operator}. The forward
operator \(\fwd\) is assumed linear and strongly smoothing (the heat
operator of the 1D heat equation introduced in Section
\ref{section:intro} is a prime example). Take a Gaussian prior
\(\param \sim \pr = \normal(\prmean ,\prcov)\) with some appropriate
covariance operator \(\prcov\) on \(\hilp\) \cite{Stuart10}. Note that
\(\fwd \prcov \fwd^*\) is the prior covariance in \(\hilo\)
\cite{Stuart10}, and as such, \(\fwd \prcov \fwd^*\) is assumed
invertible --- an assumption which will be used below (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether). Measurements are taken via a \emph{measurement operator}
\(\obs\). It is common for the measurement and forward operators to be
merged \(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but
the analysis carried out in the following sections requires that
\(\fwd\) and \(\obs\) are explicitly separated as in
\cite{attia2022stochastic, cvetkovic2023choosing}.  \(\obs \in (
\hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say \(\obs^{\star}\) is \emph{D-optimal} if \(\obs^{\star} =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let \(\hilp\) and \(\hilo\) separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively), and let
\(\fwd: \hilp \to \hilo\) the \emph{forward operator}. The forward
operator \(\fwd\) is assumed linear with fast decaying singular values
(the heat operator of the 1D heat equation introduced in Section
\ref{section:intro} is a prime example). Take a Gaussian prior
\(\param \sim \pr = \normal(\prmean ,\prcov)\) with some appropriate
covariance operator \(\prcov\) on \(\hilp\) \cite{Stuart10}. Note that
\(\fwd \prcov \fwd^*\) is the prior covariance in \(\hilo\)
\cite{Stuart10}, and as such, \(\fwd \prcov \fwd^*\) is assumed
invertible --- an assumption which will be used below (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether). Measurements are taken via a \emph{measurement operator}
\(\obs\). It is common for the measurement and forward operators to be
merged \(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but
the analysis carried out in the following sections requires that
\(\fwd\) and \(\obs\) are explicitly separated as in
\cite{attia2022stochastic, cvetkovic2023choosing}.  \(\obs \in (
\hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

%% It is important to note that since \(\obs\) is finite-rank,
%% \(\prcov^{1/2} \fwd ^* \obs^* \Sigma^{-1} \obs \fwd \prcov^{1/2}\) is
%% trace-class.
\begin{definition}\label{def:d optimality}
  We say \(\opt\) is \emph{D-optimal} if \(\opt =
  \argmax_{\obs} \tar(\obs)\), where entries of \(\obs \in (\hilo^*)^m\)
  are constrained to some allowed set of measurements in \(\hilo^*\).
\end{definition}
